---
format: gfm
execute:
  echo: false
  fig-dpi: 300
---

[![DOI](https://img.shields.io/badge/DOI-10.1038/s44298--025--00121--0-green)](https://doi.org/10.1038/s44298-025-00121-0)
[![Data](https://img.shields.io/badge/Data-10.6084/m9.figshare.27679509-green)](https://doi.org/10.6084/m9.figshare.27679509)
[![Preprint](https://img.shields.io/badge/Preprint-10.1101/2024.11.19.624162-blue)](https://www.biorxiv.org/content/10.1101/2024.11.19.624162v1)

```{r setup}
#| include: false

# packages used for data wrangling
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)

# packages used for cluster analysis
library(mice)
library(clValid)
library(NbClust)

# packages used for data visualization
library(ggplot2)
library(RColorBrewer)
library(cowplot)
```

## Enhanced encephalitic tropism of bovine H5N1 compared to the Vietnam H5N1 isolate in mice

This repository contains the data and code for the PCA analysis discussed in the manuscript, "Enhanced encephalitic tropism of bovine H5N1 compared to the Vietnam H5N1 isolate in mice". The data will be available on [Figshare](https://doi.org/10.6084/m9.figshare.27679509) after publication.

```{r data wrangling}
#| include: false

### download data ###
# -- to do --


### import data ###
dat <- read_excel("2024-08-28_H5N1data_Compiled.xlsx", skip = 1) |>
  rename(Day = `Day of harvest`,
         Weight = `Weight at euthanasia (%)`,
         Nasal_titer = `Nasal turbinates`,
         Lung_titer = Lung,
         Brain_titer = Brain)


# clean up column names for cytokine protein levels
proteins <- c('CXCL1', 'IFN-y', 'IL-1B', 'IL-2', 'IL-4', 'IL-5', 'IL-6', 'IL-10', 'IL-12p70', 'TNFa')

for(p in proteins)
{
  names(dat)[starts_with(p, vars = names(dat))][1:3] <- paste0(p, "_prot_", c('brain', 'nasal', 'lung'))
}


# clean up column names for cytokine mRNA levels
mrnas <- c('IFNa', 'IFNb', 'IFNg', 'IFNl2', 'IL1b', 'IL6', 'TNFa', 'ISG15', 'ISG56', 'Mx1', 'OAS1')

for(m in mrnas)
{
  names(dat)[starts_with(paste0(m, '.'), vars = names(dat))] <- paste0(m, "_mRNA_", c('brain', 'nasal', 'lung'))
}


### convert to long format ###

# variables to convert column names so that we are working with the same column name formats for everything
name_conversion <- tibble(colnumbers = starts_with(c('IFNa', 'IFNb', 'IFNg', 'IFNl2', 'IL1b', 'IL6', 'TNFa', 'ISG15', 'ISG56', 'Mx1', 'OAS1', 'CXCL1', 'IFN-y', 'IL-1B', 'IL-2', 'IL-4', 'IL-5', 'IL-6', 'IL-10', 'IL-12p70', 'TNFa'), vars = names(dat)),
                          colname = names(dat)[colnumbers],
                          name_parts = str_split(colname, "_"),
                          tissue = map_chr(name_parts, ~.x[3]),
                          chemokine = map_chr(name_parts, ~.x[1]),
                          type = map_chr(name_parts, ~.x[2]),
                          new_colname = paste(tissue, chemokine, type, sep = "_"))
convert_names <- name_conversion$new_colname
names(convert_names) <- name_conversion$colname


dat_long <- dat |>
  plyr::rename(convert_names) |>

  # this doesn't fit into the PCA comparing the different tissues
  select(-Weight) |>

  # make super long (we will get a warning for type in titer - this is OK)
  pivot_longer(cols = -c(1:5),
               names_sep = "_",
               names_to = c("tissue", "chemokine", "type"),
               values_to = "value") |>

  mutate(tissue = tolower(tissue)) |>

  # widen back out some for PCA with tissue as a factor
  pivot_wider(names_from = c("chemokine", "type"), values_from = value) |>

  rename(titer = titer_NA)


# fix some column names
names(dat_long) <- gsub("-", "", names(dat_long))


# log10 transform titer
for(i in 7:dim(dat_long)[2])
{
  dat_long[[i]] <- log10(dat_long[[i]])
}
```

```{r pca}
#| include: false

### missingness and variability checks ###

# these will flag columns we want to drop for PCA
# (not including categorical data in columns 1:6)
drop_day3 <- c(rep(TRUE, 6), rep(FALSE, dim(dat_long)[2] - 6))
names(drop_day3) <- names(dat_long)

drop_endpoint <- drop_day3

# we'll track for day 3 and the endpoint separately
day3 <- dat_long$Day == 3

for(j in 7:dim(dat_long)[2])
{
  # tabulate missingness of jth column by tissue
  miss_3   <- table(is.na(dat_long[[j]][ day3]), dat_long$tissue[ day3]) / sum( day3)
  miss_end <- table(is.na(dat_long[[j]][!day3]), dat_long$tissue[!day3]) / sum(!day3)

  # check for variability
  var_3   <- var(dat_long[[j]][ day3], na.rm = TRUE)
  var_end <- var(dat_long[[j]][!day3], na.rm = TRUE)
  
  # check if any of the tissues has > 20% missingness or is essentially constant
  drop_day3    [j] <- any(miss_3  [-1,] > 0.2) | (var_3   < 1e-6)
  drop_endpoint[j] <- any(miss_end[-1,] > 0.2) | (var_end < 1e-6)
}


### PCA ###

set.seed(238974)
dat_pca_3 <- filter(dat_long, Day == 3)[,!drop_day3] |>

  mice() |>                                      # impute missing values
  complete() |>

  princomp(cor = TRUE)                           # PCA

pca_results_3 <- cbind(dat_long[day3, 1:6], dat_pca_3$scores)
pca_pct_variability_3 <- cumsum(dat_pca_3$sdev) / sum(dat_pca_3$sdev) * 100

dat_pca_end <- filter(dat_long, Day > 3)[,!drop_day3] |>

  mice() |>                                      # impute missing values
  complete() |>

  princomp(cor = TRUE)                           # PCA

pca_results_end <- cbind(dat_long[!day3, 1:6], dat_pca_end$scores)
pca_pct_variability_end <- cumsum(dat_pca_end$sdev) / sum(dat_pca_end$sdev) * 100


### clustering ###

if(FALSE) # only run this part manually - not explicitly included in README
{
  # check optimal number of clusters
  clValid(dat_pca_3$scores, 2:10, clMethods = c("kmeans"), validation = "internal") |>
    plot()
  
  clValid(dat_pca_end$scores, 2:10, clMethods = c("kmeans"), validation = "internal") |>
    plot()
  
  # suggests 2 clusters for Day 3
  nb_day3 <- NbClust(dat_pca_3$scores, distance = "euclidean", 
                     min.nc = 2, max.nc = 10, method = "kmeans")

  # suggests 3 clusters for endpoint
  nb_end <- NbClust(dat_pca_end$scores, distance = "euclidean",
                    min.nc = 2, max.nc = 10, method = "kmeans") 
}

# add cluster information to the data
pca_results_3$cluster <- kmeans(dat_pca_3$scores, centers = 2)$cluster
pca_results_end$cluster <- kmeans(dat_pca_end$scores, centers = 3)$cluster
```

The Principal Component Analysis (PCA) of viral titer, protein abundance and mRNA levels in brain, lung and nasal turbinates of mice, separated by day of harvest (day 3 and endpoint). The first two principal components accounted for `r round(pca_pct_variability_3[2])`% and `r round(pca_pct_variability_end[2])`% of total sample variability in day 3 and endpoint samples, respectively.

Analysis at 3 days post inoculation resulted in two distinct clusters: the smaller cluster contained only lung samples, all but one of which were high dose groups. The remaining low-dose lung samples, all brain, and all nasal samples were contained in the larger cluster.

```{r}
# one figure to rule them all!
make_pca_plot <- function(pca_results, lbl)
{
  hulls <- pca_results |>
    group_by(cluster) |>
    summarise(hull = list(chull(Comp.1, Comp.2))) |>
    pull(hull)

  # base plot
  pca_plt <- mutate(pca_results,
                    Group = paste(Virus, Dose)) |>
    ggplot(aes(x = Comp.1, y = Comp.2, color = Group, shape = tissue))

  for(i in 1:length(hulls)) {
    pca_plt <- pca_plt +
      geom_polygon(data = filter(pca_results, cluster == i)[hulls[[i]],],
                   aes(Comp.1, Comp.2, group = cluster),
                   fill = 'grey97', color = 'black', linewidth = 0.1)
  }

  pca_plt +
    geom_point() +
    labs(x = paste0("PC 1 (", lbl, ")"),
         y = paste0("PC 2 (", lbl, ")"),
         shape = 'Tissue') +
    scale_color_brewer(palette = "Paired",
                       labels = expression(paste('Bovine ', 10^2),
                                           paste('Bovine ', 10^4),
                                           paste('VN1203 ', 10^2),
                                           paste('VN1203 ', 10^4))) +
    theme_cowplot() +
    theme(plot.title = element_text(hjust = -0.2))
}

make_pca_plot(pca_results_3, 'Day 3')
```

In the PCA of samples collected at endpoint, the optimal number of clusters was determined to be three, with clusters similar in size. One cluster was primarily composed of VN1203-exposed samples, while a second cluster mainly contained Bov342-exposed samples, both predominantly from brain and nasal turbinate samples. The third cluster contained only lung samples, where all but one of the VN1203exposed mice clustered together with most of the high-dose Bov342-exposed mice. Most of the low-dose Bov342-exposed lung tissues clustered with the Bov342-exposed brain and nasal samples.

```{r}
make_pca_plot(pca_results_end, 'Endpoint')
```
